---
title: ""From pandemic to endemic: divergence of COVID-19 positive-tests and hospitilization numbers from SARS-CoV-2 RNA levels in wastewater of Rochester, Minnesota"
author: "Dr. Biruhalem Taye"
date: "04/07/2023"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```


#**Rochester, Byron and Stewartville SARS-CoV-2 in wastewater and new COVID-19 cases trend and correlation analysis** 

```{r}
#loading R libraries 
library(dplyr)
library(tidyverse)
library(reticulate)
library(rstatix)
library(ggpubr)
library(ggplot2)

```


*loading and formatting data*

```{r}
#loading data

alldata <- read.csv("C:/Users/biruh/Desktop/Wastwater/Figuresversion2_V2/2023Alldatarefined.csv")

#head(alldata) 

#Converting character data to date to be recognized by R 

alldata$Days <- as.Date(alldata$Days,"%m/%d/%Y")

alldata$Weekday <- format(alldata$Days, '%A') # annotating weekdays 

#head(alldata) #inspecting the data 


```


#Plotting the trend data using python malplotlib.pyplot. We use reticulate R package to integrate R and python scripts



```{python}
#importing python libraries 

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
import matplotlib.ticker as ticker
import matplotlib.dates as mdates
from matplotlib.dates import (AutoDateLocator, YearLocator, MonthLocator,
                              DayLocator, WeekdayLocator, HourLocator,
                              DateFormatter,
                              AutoDateFormatter, ConciseDateFormatter)

```

#Rochester trend analysis (Fig 3A)

```{python}
# Creating plot with cases
fig, ax1 = plt.subplots(figsize=(12, 8))


color = 'tab:red'
#ax1.set_xlabel('Weeks')
ax1.set_ylabel('SARS-CoV-2 in wastewater', fontsize=20, fontweight = "bold", color = color)
ax1.scatter(r.alldata.Days, r.alldata.Rochester_SARS, color = color)
ax1.tick_params(axis ='y', labelcolor = color, labelsize = 18)


# Adding Twin Axes to plot using dataset_2
ax2 = ax1.twinx()
 
color = 'tab:blue'
ax2.set_ylabel('Daily COVID-19 cases', fontsize=20, fontweight ="bold", color = color)
ax2.bar(r.alldata.Days, r.alldata.Rochester_Cases, color = color)
#ax1.xaxis.set_major_locator(mdates.ConciseDateFormatter())
#ax1.xaxis.set_major_locator(mdates.MonthLocator())
#ax1.xaxis.set_major_formatter(mdates.DateFormatter('%b %Y'))
ax2.tick_params(axis = 'x', rotation=30, labelsize = 16)
#ax1.set_tick_params(axis = 'x', weight = 'bold')
ax2.tick_params(axis ='y', labelcolor = color, labelsize = 18)
ax2.axvline(pd.to_datetime('2022-04-25'), color='g', linestyle='--', lw=1)
ax2.axvline(pd.to_datetime('2022-09-15'), color='r', linestyle='--', lw=1) # change point 


 
# Adding title

#plt.title('Rochester daily COVID-19 cases vs SARS-CoV-2 in Wastwater ', fontweight ="bold")

plt.savefig('Rochester_daily_bar_scatter_invert.png') 
# Show plot
plt.show()

```


#Byron trend analysis (Fig 3B)

```{python}
# Creating plot with cases
fig, ax1 = plt.subplots(figsize=(12, 8))


color = 'tab:red'
#ax1.set_xlabel('Weeks')
ax1.set_ylabel('SARS-CoV-2 in wastewater', fontsize=20, fontweight = "bold", color = color)
ax1.scatter(r.alldata.Days, r.alldata.Byron_SARS, color = color)
ax1.tick_params(axis ='y', labelcolor = color, labelsize = 18)


# Adding Twin Axes to plot using dataset_2
ax2 = ax1.twinx()
 
color = 'tab:blue'
ax2.set_ylabel('Daily COVID-19 cases', fontsize=20, fontweight ="bold", color = color)
ax2.bar(r.alldata.Days, r.alldata.Byron_Cases, color = color)
#ax1.xaxis.set_major_locator(mdates.ConciseDateFormatter())
#ax1.xaxis.set_major_locator(mdates.MonthLocator())
#ax1.xaxis.set_major_formatter(mdates.DateFormatter('%b %Y'))
ax2.tick_params(axis = 'x', rotation=30, labelsize = 16)
#ax1.set_tick_params(axis = 'x', weight = 'bold')
ax2.tick_params(axis ='y', labelcolor = color, labelsize = 18)
ax2.axvline(pd.to_datetime('2022-04-25'), color='g', linestyle='--', lw=1)
ax2.axvline(pd.to_datetime('2022-09-27'), color='r', linestyle='--', lw=1) # change point 


 
# Adding title

#plt.title('Rochester daily COVID-19 cases vs SARS-CoV-2 in Wastwater ', fontweight ="bold")

plt.savefig('Byron_daily_bar_scatter_invert.png') 
# Show plot
plt.show()

```


#Stewartville trend analysis (Fig 3C)

```{python}
# Creating plot with cases
fig, ax1 = plt.subplots(figsize=(12, 8))


color = 'tab:red'
#ax1.set_xlabel('Weeks')
ax1.set_ylabel('SARS-CoV-2 in wastewater', fontsize=20, fontweight = "bold", color = color)
ax1.scatter(r.alldata.Days, r.alldata.Stewartville_SARS, color = color)
ax1.tick_params(axis ='y', labelcolor = color, labelsize = 18)


# Adding Twin Axes to plot using dataset_2
ax2 = ax1.twinx()
 
color = 'tab:blue'
ax2.set_ylabel('Daily COVID-19 cases', fontsize=20, fontweight ="bold", color = color)
ax2.bar(r.alldata.Days, r.alldata.Stewartville_Cases, color = color)
#ax1.xaxis.set_major_locator(mdates.ConciseDateFormatter())
#ax1.xaxis.set_major_locator(mdates.MonthLocator())
#ax1.xaxis.set_major_formatter(mdates.DateFormatter('%b %Y'))
ax2.tick_params(axis = 'x', rotation=30, labelsize = 16)
#ax1.set_tick_params(axis = 'x', weight = 'bold')
ax2.tick_params(axis ='y', labelcolor = color, labelsize = 18)
ax2.axvline(pd.to_datetime('2022-04-25'), color='g', linestyle='--', lw=1)
ax2.axvline(pd.to_datetime('2022-05-31'), color='r', linestyle='--', lw=1) # change point 


 
# Adding title

#plt.title('Rochester daily COVID-19 cases vs SARS-CoV-2 in Wastwater ', fontweight ="bold")

plt.savefig('Srewartville_daily_bar_scatter_invert.png') 
# Show plot
plt.show()

```


#**Normality test**

Normality test (Testing for normal/Gaussian distribution of the data) 

Shapiro-Wilk normality test: 

If the p-values > 0.05 the data is normally distributed. P < 0.05 the data doesn't assume normal distribution. 


```{r}
library(rstatix)

shapiro_test(alldata, Rochester_N1, Rochester_N2, Rochester_SARS, Rochester_Cases, Stewartville_SARS, Stewartville_Cases, Byron_SARS, Byron_Cases)

```

Based on Shapiro-Wilk normality test all the data are not normally distributed, hence Spearman correlation is used for correlation analysis.

**Spearman's rank correlations**

**N1 and N2 target sequences**

Correlation of SARS-CoV-2 in wastewater using nucleocapsid target sequences N1 and N2 

#Supplementary Figure 2

```{r}
p1_n <- ggscatter(alldata, x = "Rochester_N1", y = "Rochester_N2", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Viral RNA per ml (N1)", ylab = "Viral RNA per ml (N2)",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

figS1 <- p1_n + theme(axis.title = element_text(size = 24),
             axis.text = element_text(size = 24))
figS1

```


**Rochester SARS-CoV-2 and COVID-19 cases correlation: Overall**  

#Fig 5A

```{r}
Rall <- ggscatter(alldata, x = "Rochester_Cases", y = "Rochester_SARS", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Daily positive cases", ylab = "SARS-CoV-2 in wastewater",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

Rall_1 <- Rall + theme(axis.title = element_text(size = 22),
             axis.text = element_text(size = 22))
Rall_1
```
 
**Rochester data correlation before change in case reporting guideline 2022-04-25** 
# Fig 5B

```{r}

alldata_befor <- alldata %>% filter(Days <= '2022-04-25')
#tail(alldata_befor)

#selection of relevant variables 
#wdata = selected relevent dataset 

Rall_b <- ggscatter(alldata_befor, x = "Rochester_Cases", y = "Rochester_SARS", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Daily positive cases", ylab = "SARS-CoV-2 in wastewater",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

Rall_b1 <- Rall_b + theme(axis.title = element_text(size = 22),
             axis.text = element_text(size = 22))
Rall_b1

```

**Rochester data correlation after 2022-04-25**

#Fig 5C

```{r}
alldata_after <- alldata %>% filter(Days > '2022-04-25')
#tail(alldata_after)

#selection of relevant variables 
#wdata = selected relevent dataset 

Rall_a <- ggscatter(alldata_after, x = "Rochester_Cases", y = "Rochester_SARS", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Daily positive cases", ylab = "SARS-CoV-2 in wastewater",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

Rall_a1 <- Rall_a + theme(axis.title = element_text(size = 22),
             axis.text = element_text(size = 22))
Rall_a1
```


**Byron data correlation overall (after 2021-06-15)**

#Supplementary Figure 4A

```{r}
Byron_all <- alldata %>% filter(Days >= '2021-06-15')

head(Byron_all)


Byall <- ggscatter(Byron_all, x = "Byron_Cases", y = "Byron_SARS", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Daily positive cases", ylab = "SARS-CoV-2 in wastewater",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

Byall_1 <- Byall + theme(axis.title = element_text(size = 22),
             axis.text = element_text(size = 22))
Byall_1

```


**Byron data correlation before 2022-04-25**

#Supplementary Figure 4B

```{r}
Byron_before <- alldata %>% filter(dplyr::between(Days, as.Date('2021-06-15'), as.Date('2022-04-25')))

#head(Byron_before)


Bybefore <- ggscatter(Byron_before, x = "Byron_Cases", y = "Byron_SARS", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Daily positive cases", ylab = "SARS-CoV-2 in wastewater",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

Bybefore1 <- Bybefore + theme(axis.title = element_text(size = 22),
             axis.text = element_text(size = 22))
Bybefore1

```

**Byron data correlation after 2022-04-25**

#Supplementary Figure 4C

```{r}
Byron_after <- alldata %>% filter(Days > '2022-04-25')

#head(Byron_after)


Byafter <- ggscatter(Byron_after, x = "Byron_Cases", y = "Byron_SARS", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Daily positive cases", ylab = "SARS-CoV-2 in wastewater",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

Byafter1 <- Byafter + theme(axis.title = element_text(size = 22),
             axis.text = element_text(size = 22))
Byafter1

```


**Stewartville overall data correlation (after 2021-05-25)**

#Supplementary Figure 5A


```{r}
stew_all <- alldata %>% filter(Days >= '2021-05-25')

#head(stew_all)


Stewall <- ggscatter(stew_all, x = "Stewartville_Cases", y = "Stewartville_SARS", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Daily positive cases", ylab = "SARS-CoV-2 in wastewater",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

Stewall1 <- Stewall + theme(axis.title = element_text(size = 22),
             axis.text = element_text(size = 22))
Stewall1

```


**Stewartville data correlation before 2022-04-25 and (after 2021-05-25)**

#Supplementary Figure 5B

```{r}
stew_before <- alldata %>% filter(dplyr::between(Days, as.Date('2021-05-25'), as.Date('2022-04-25')))

Stewbefore <- ggscatter(stew_before, x = "Stewartville_Cases", y = "Stewartville_SARS", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Daily positive cases", ylab = "SARS-CoV-2 in wastewater",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

Stewbefore <- Stewbefore + theme(axis.title = element_text(size = 22),
             axis.text = element_text(size = 22))
Stewbefore

```

**Stewartville data correlation after 2022-04-25** 

#Supplementary Figure 5C


```{r}
stew_after <- alldata %>% filter(Days >= '2022-04-25')

#head(stew_after)


stewafter <- ggscatter(stew_after, x = "Stewartville_Cases", y = "Stewartville_SARS", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Daily positive cases", ylab = "SARS-CoV-2 in wastewater",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

stewafter <- stewafter + theme(axis.title = element_text(size = 22),
             axis.text = element_text(size = 22))
stewafter
```

**SARS-CoV-2 concentration in wastewater and COVID-19 hospitalizations**

```{r}
#loading data 
alldata_hos <- read.csv("C:/Users/biruh/Desktop/Wastwater/Figuresversion2_v2/2023Alldatarefined_hostpitalization.csv")

head(alldata_hos) 

#Converting character data to date to be recognized by R 

alldata_hos$Days <- as.Date(alldata$Days,"%m/%d/%Y")

library(dplyr)
alldata_hos$Weekday <- format(alldata$Days, '%A')

head(alldata_hos)
```


Rochester trend SARS-CoV-2 in wastewater vs Hospitalization

#Fig 6A

```{python}
# Creating plot with cases
fig, ax1 = plt.subplots(figsize=(12, 8))


color = 'tab:red'
#ax1.set_xlabel('Weeks')
ax1.set_ylabel('SARS-CoV-2 in wastewater', fontsize=20, fontweight = "bold", color = color)
ax1.scatter(r.alldata_hos.Days, r.alldata_hos.Rochester_SARS, color = color)
ax1.tick_params(axis ='y', labelcolor = color, labelsize = 18)


# Adding Twin Axes to plot using dataset_2
ax2 = ax1.twinx()
 
color = 'k'
ax2.set_ylabel('Daily COVID-19 hospitalization', fontsize=20, fontweight ="bold", color = color)
ax2.bar(r.alldata_hos.Days, r.alldata_hos.Rochester_NewHosp, color = color)
#ax1.xaxis.set_major_locator(mdates.ConciseDateFormatter())
#ax1.xaxis.set_major_locator(mdates.MonthLocator())
#ax1.xaxis.set_major_formatter(mdates.DateFormatter('%b %Y'))
ax2.tick_params(axis = 'x', rotation=30, labelsize = 16)
#ax1.set_tick_params(axis = 'x', weight = 'bold')
ax2.tick_params(axis ='y', labelcolor = color, labelsize = 18)
ax2.axvline(pd.to_datetime('2022-04-25'), color='g', linestyle='--', lw=1)
ax2.axvline(pd.to_datetime('2022-09-15'), color='r', linestyle='--', lw=1) # change point 


 
# Adding title

#plt.title('Rochester daily COVID-19 hospitalization vs SARS-CoV-2 in Wastwater ', fontweight ="bold")

plt.savefig('Rochester_daily_bar_scatter_host_invert.png') 
# Show plot
plt.show()
```


#Correlation of SARS-CoV-2 in wastewater vs Rochester Hospitilization overall

```{r}
Rall_hosp_all <- ggscatter(alldata_hos, x = "Rochester_NewHosp", y = "Rochester_SARS", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Daily COVID-19 hospitilization", ylab = "SARS-CoV-2 in wastewater",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

Rall_hosp_all1 <- Rall_hosp_all + theme(axis.title = element_text(size = 22),
             axis.text = element_text(size = 22))
Rall_hosp_all1

```

#Correlation of SARS-CoV-2 in wastewater vs Rochester Hospitilization before 2022-04-25

#Fig 6B

```{r}
alldata_hos_before <- alldata_hos %>% filter(Days <= '2022-04-25')
tail(alldata_hos_before)

Rall_hosp <- ggscatter(alldata_hos_before, x = "Rochester_NewHosp", y = "Rochester_SARS", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Daily COVID-19 hospitilization", ylab = "SARS-CoV-2 in wastewater",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

Rall_hosp1 <- Rall_hosp + theme(axis.title = element_text(size = 22),
             axis.text = element_text(size = 22))
Rall_hosp1


```

#Correlation of SARS-CoV-2 in wastewater vs Rochester Hospitilization after 2022-04-25

#Fig 6C

```{r}
alldata_hos_after <- alldata_hos %>% filter(Days > '2022-04-25')
tail(alldata_hos_after)

Rall_hosp_a <- ggscatter(alldata_hos_after, x = "Rochester_NewHosp", y = "Rochester_SARS", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          add.params = list(color = "blue", fill = "lightgray"),
          xlab = "Daily COVID-19 hospitilization", ylab = "SARS-CoV-2 in wastewater",
          size = 2,
          cor.coef.size = 8,
          cor.coef.font = c(24, "bold", "black"),
          pvalue.font = c(24, "bold", "black"))

Rall_hosp2 <- Rall_hosp_a + theme(axis.title = element_text(size = 22),
             axis.text = element_text(size = 22))
Rall_hosp2

```


**Modeling**

Loading libraries (R-packages)

```{r}
#Loading libraries 

library(dLagM)
library(tictoc)
library(lmtest)
library(tseries)
library(forecast)
library(pracma)
library(dlnm)

```


#loading interpolated Rochester data 

```{r}

roch <- read.csv("C:/Users/biruh/Desktop/Wastwater/2023rochesterdata2.csv")

#Converting character data to date to be recognized by R 

roch$Days <- as.Date(roch$Days,"%m/%d/%Y")

#Creating column that corresponds the days from Monday to Sunday.
roch$nameoftheday <- format(roch$Days, '%A')

wdata <- roch %>% select(1,9,10, 11) # selection of column 1, 9, 10 and 11 (relevant data)


```

Extracting data before 2022-04-25

```{r}
#pcr = extracted data before 2022-04-25 (440 days)

pcr <- wdata %>% filter(Days <= '2022-04-25')
#head(pcr)

```


**Modeling**

For all modeling analysis we used the data before changes in guideline in COVID-19 case report (data before 2022-04-25 (440 days))



**Distributed lag models and autoregressive distributed lag (ARDL)**



Distributed lag model was selected based on previous publication.

https://pubmed.ncbi.nlm.nih.gov/36380770/

An R package "dLagM" and its dependencies were used for analysis

(https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0228812)


A distributed lag model can be an appropriate choice when there is a lagged relationship between the independent variable and the dependent variable. 

independent variable = N1_N2gflowadj,

dependent variable = Daily_cases

**MODEL 1**

**Finite distributed lag model**

```{r}
dlmFit1 <- dlm(x = pcr$N1_N2gflowadj, y = pcr$Daily_cases, q = 15)
summary(dlmFit1)

```

**MODEL 2**

**Autoregressive Distributed Lag (ARDL) model** (with p = 15)


```{r}

ARDLfit2 <- ardlDlm(x = pcr$N1_N2gflowadj, y = pcr$Daily_cases, p=15, q=15)
summary(ARDLfit2)

```


**MODEL 3**

**ARIMA Model**

```{r}

ARIMA <- auto.arima(pcr$Daily_cases, xreg = pcr$N1_N2gflowadj)
summary(ARIMA)


checkresiduals(ARIMA)

accuracy(ARIMA)
```




**GOODNESS-OF-FIT MEASURES**


```{r}

sortScore(x = MASE(dlmFit1, ARDLfit2), score = c("mase"))

```


```{r}
bestfitted_model <- GoF(dlmFit1, ARDLfit2)
print(bestfitted_model)
```







