---
title: "From pandemic to endemic: summer 2022 divergence of COVID-19 case numbers and SARS-CoV-2 RNA detection in wastewaters of Rochester, Minnesota"
author: "Dr Biruhalem Taye"
date: "04/07/2023"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```


#**Supplementary data** 

```{r}
#loading R libraries 
library(dplyr)
library(tidyverse)
library(reticulate)
library(rstatix)
library(ggpubr)
library(ggplot2)

```


*loading and formatting data*

```{r}
#loading data

alldata <- read.csv("C:/Users/biruh/Desktop/Wastwater/Figuresversion2/2023Alldatarefined.csv")

#head(alldata) 

#Converting character data to date to be recognized by R 

alldata$Days <- as.Date(alldata$Days,"%m/%d/%Y")

alldata$Weekday <- format(alldata$Days, '%A') # annotating weekdays 

#head(alldata) #inspecting the data 


```


**Normality test**

Normality test (Testing for normal/Gaussian distribution of the data) 

Shapiro-Wilk normality test: 

If the p-values > 0.05 the data is normally distributed. P < 0.05 the data doesn't assume normal distribution. 


```{r}
library(rstatix)

shapiro_test(alldata, Rochester_N1, Rochester_N2, Rochester_SARS, Rochester_Cases, Stewartville_SARS, Stewartville_Cases, Byron_SARS, Byron_Cases)

```

Based on Shapiro-Wilk normality test all the data are not normally distributed, hence Spearman correlation is used for correlation analysis.

**Modeling**

Loading libraries (R-packages)

```{r}
#Loading libraries 

library(dLagM)
library(tictoc)
library(lmtest)
library(tseries)
library(forecast)
library(pracma)
library(dlnm)

```


#loading interpolated Rochester data 

```{r}

roch <- read.csv("C:/Users/biruh/Desktop/Wastwater/2023rochesterdata2.csv")

#Converting character data to date to be recognized by R 

roch$Days <- as.Date(roch$Days,"%m/%d/%Y")

#Creating column that corresponds the days from Monday to Sunday.
roch$nameoftheday <- format(roch$Days, '%A')

wdata <- roch %>% select(1,9,10, 11) # selection of column 1, 9, 10 and 11 (relevant data)


```

Extracting data before 2022-04-25

```{r}
#pcr = extracted data before 2022-04-25 (440 days)

pcr <- wdata %>% filter(Days <= '2022-04-25')
#head(pcr)

```


**Modeling**

For all modeling analysis we used the data before changes in guideline in COVID-19 case report (data before 2022-04-25 (440 days))



**Distributed lag models and autoregressive distributed lag (ARDL)**



Distributed lag model was selected based on previous publication.

https://pubmed.ncbi.nlm.nih.gov/36380770/

An R package "dLagM" and its dependencies were used for analysis

(https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0228812)


A distributed lag model can be an appropriate choice when there is a lagged relationship between the independent variable and the dependent variable. 

independent variable = N1_N2gflowadj,

dependent variable = Daily_cases

**MODEL 1**

**Finite distributed lag model**

```{r}
dlmFit1 <- dlm(x = pcr$N1_N2gflowadj, y = pcr$Daily_cases, q = 15)
summary(dlmFit1)

```

**MODEL 2**

**Autoregressive Distributed Lag (ARDL) model** (with p = 8)


```{r}

ARDLfit2 <- ardlDlm(x = pcr$N1_N2gflowadj, y = pcr$Daily_cases, p=8, q=15)
summary(ARDLfit2)

```


**GOODNESS-OF-FIT MEASURES**


```{r}

sortScore(x = MASE(dlmFit1, ARDLfit2), score = c("mase"))

```


```{r}
bestfitted_model <- GoF(dlmFit1, ARDLfit2)
print(bestfitted_model)
```







